---
title: "Sonify data - Intro"
author: "mark johnson"
date: "11/21/2016"
output: html_document
---

## Sonify data in R
Why? Good question!  

How? workflow as follows:  
1. Set up with necessary packages and WavPlayer  
2. Ingest a time series environmental dataset   
3. Massage as needed  
3.1 Eliminate NAs  
3.2 Assess object length relative to how long you the audio representation of it to last  
4. Sonify  

```{r}
# 1.0 load necessary R packages
library(seewave)
library(tuneR)
library(zoo) # needed for some gap filling in a worked example 
setWavPlayer("afplay") # other options include "mpv", VLC, quicktime, etc. 
# setWavPlayer("/Applications/VLC.app/Contents/MacOS/VLC") 
# Above line with VLC is untested, but may work if path makes sense and you have VLC installed
```

Next, load up your time series of choice. Here, we have two examples, one that needs some massaging, one that doesn't. 

```{r}
# 2.0 Load in a time series of environmental data. Two examples follow
CO2_data <- read.table("ftp://aftp.cmdl.noaa.gov/products/trends/co2/co2_mm_mlo.txt",
             sep = "", row.names = NULL,header = F,colClasses = rep("numeric", 7),
             comment.char = "#", na.strings = -99.99, 
             col.names = c("Yr", "Mo", "Mo_Yr", "CO2", "Trnd", "X", "dys"))
head(CO2_data)

# next example is water level measurements in a flooded forest
pantanal <- read.csv("http://ecohydro.ires.ubc.ca/pdfs/CZ.waterlevel.csv")
head(pantanal)
```

Proceed with any necessary data massaging/wrangling. NAs are likely problematic. Gap-fill as appropriate. 

```{r}
# 3.0 Wrangle your data
# gap-fill to replace the NA's in the CO2 data
CO2_data$CO2.gf <- na.locf(CO2_data$CO2)
head(CO2_data) #check that no NAs in gap-filled data (e.g. CO2_data$CO2.df)
# Now, the CO2_data$CO2.gf is ready to sonify (as is pantanal$Dw.TT)
```

Test out your speakers and listen 

```{r}
# sheep sound
data(sheep)
listen(sheep,f=8000)
str(sheep)
plot(sheep@left)
oscillo(sheep)

# bird sound
data(tico)
listen(tico,f=8000)
oscillo(tico)
```

Give it a try with the provided data or your own. 
Below we map the variable range onto an audible frequency range. 

```{r}
# select a vector to sonify (this will be called myVar)
#myVar <- pantanal$Dw.TT # uncomment if want the water level data for a flooded forest
myVar <- CO2_data$CO2.gf
zVar <- (myVar - mean(myVar)) / sd(myVar)  #standardize data to mean = 0 and sd = 1 (though not really needed)

# map the variable range onto an audible frequency range
f.low <- 200 # frequency of lowest pitch
f.high <- 4000 # frequency of highest pitch
m.sound <- (f.high - f.low) / (max(myVar) - min(myVar)) #slope of the data-to-sound relationship
b.sound <- (f.high + f.low - (max(myVar)+min(myVar))*m.sound)/2   # intercept of the data-to-sound relationship

# expand the data set (if needed) using the rep function. Suggested value of 100 for CO2_data and 10 for Pantanal water level. Adjust as desired. 
myVar.expanded <- rep(myVar, each = 100)  # the longer the "each", the longer the note for each data point

myVar.freq <- myVar.expanded * m.sound + b.sound

# select harmonics
clarinet <- c(1, 0, 0.5, 0, 0.14, 0, 0.5, 0, 0.12, 0, 0.17) 
s <- synth2(ifreq=myVar.freq, f=8000, harmonics = clarinet, listen = FALSE, plot = TRUE, output = "Wave")
str(s)
# # object "s" is now stored as a Wave class object, and can be output to use in presentations, websites, random acts of geekery, etc.
listen(s) 
```

functionalizing

```{r}
# make a function of the data-to-sound mapping
data2sound <- function(x, freq.low, freq.high, expansion.factor){
  b <- coef(lm(c(freq.low, freq.high) ~ range(x)))[1]
  m <- coef(lm(c(freq.low, freq.high) ~ range(x)))[2]
  x.freq <- m * x + b
  return(x.freq.expand <- rep(x.freq, each = expansion.factor))
  }
```

test the function data2sound

```{r}
myVar <- pantanal$Dw.TT # water level data for a flooded forest

tmp.var <- data2sound(myVar,200,4000,10)
s2 <- synth2(ifreq = tmp.var, f=8000, harmonics = clarinet, listen = FALSE, output = "Wave")
listen(s2)
```

Next steps: 
Get back to work ...or... build this into a Shiny app
